---
title: "ACS Census Data"
output:
  html_document:
    df_print: paged
---

Code to examine the ACS Census data, which contains statistics about gender and pay as well. This will be based on the tidycensus package (https://github.com/GL-Li/totalcensus)

## Set Up Our Code

We need to load in the correct packages, and set up an API key and a path where we want our data to be loaded to. The Census API offers an easy way to download data. You need to get a key from this website, http://api.census.gov/data/key_signup.html. Once you have a key, you can replace the text and use replace "YOUR API KEY GOES HERE"

```{r setup, echo=FALSE, message=FALSE}
# install.packages(c("tidycensus","tidyverse", "totalcensus","ggplot2"))
# install.packages(c("maptools","tigris", "leaflet","ggmap","totalcensus"))
library(tidycensus)
library(tidyverse)
library(totalcensus)
library(ggplot2)
library(ggmap)
library(maptools)
library(tigris) # for function readOGR to get geo ID and polygon squares
library(leaflet)
#census_api_key("YOUR API KEY GOES HERE")
#census_api_key("71b73631ae7447defddb4c42e85a20e970e2c7bb", install=TRUE)
census_api_key("71b73631ae7447defddb4c42e85a20e970e2c7bb")
```

We want to examine the different options associated with the ACS data. Feel free to filter the different datasets below to generate different data.

```{r info, echo=TRUE}
vars <- search_tablecontents("acs", view=FALSE) #figure out which variables you want
summary_levs <- search_summarylevels("acs", view=FALSE) # figure out which summary levels we have access to
geos <- search_geocomponents("acs", view=FALSE)
geos2 <- search_geoheaders("acs", view=FALSE)
```

### DMV Map

We want to create the first base map of the DMV, or DC-Maryland-Virginia area. We'll use county level data.  

```{r mapdat, echo=TRUE, results='hide'}
#from the tigris package we can get each individual county from state
states <- c("DC","VA","MD")
dmv <- rbind_tigris(
  lapply(
    states, function(x) counties(state = x,cb =TRUE)
  )
)

```

```{r dmv-map, echo=FALSE}
plot(dmv) 
```




### Visualize the Gender Ratio in the DMV.

We want to visualize the gender ration in each area. We'll create a function 'load_data_fm' to get ACS 5 year county estimates for males and females, for each state.  

```{r male-female-ratios, echo=TRUE}
load_data_fm <- function(state){
  return(get_acs("county", variables=c(male="B01001_002",female="B01001_026"), state=state, survey="acs5"))
}

dc_fm <- load_data_fm("DC")
md_fm <- load_data_fm("MD")
va_fm <- load_data_fm("VA")

dmv_fm <- rbind(dc_fm, md_fm, va_fm) #combine each state together
dmv_fm <- dmv_fm[,-5]

dmv_fm2 <- dmv_fm %>% spread(variable, estimate) #transpose the male and female variables
dmv_fm2$pfemale <- round(dmv_fm2$female / (dmv_fm2$female + dmv_fm2$male) * 100 , 0) #calculate the female percentage

print(head(dmv_fm2))


```

Now we can map the percentage female in each county. 

```{r percent-female, echo=TRUE}
dmv2 <- geo_join(dmv, dmv_fm2, "GEOID", "GEOID", how="inner") #158 rows

pal <- colorQuantile("Blues", NULL, n = 3)

map1 <- leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = dmv2, 
              fillColor = ~pal(dmv2$pfemale), 
              fillOpacity = 0.7, 
              weight = 0.2, 
              smoothFactor = 0.2, 
              popup = paste("Percent Female: ", dmv2$pfemale, "%", sep="" )) %>%
  addLegend(pal = pal, 
            values = dmv2$pfemale, 
            position = "bottomright", 
            title = "Percent Female")

map1
```

### Gender Inequality Statistics

We want to filter for the counts of individuals in each category. The table_contents variable is how we do that. m1_2499 translates to Male $1-$2,499 income level count. 

```{r loadpaydata, echo=TRUE}

load_data_pay <- function(state){
  return(get_acs("county", variables=c(m1_2499 = "B20001_003", m2500_4999 = "B20001_004",
                                       m5000_7499 ="B20001_005",
                       m7500_9999 = "B20001_006",m10000_12499 = "B20001_007",
                       m12500_14999 = "B20001_008",m15000_17499 =  "B20001_009",
                       m17500_19999 = "B20001_010",
                       m20000_22499 = "B20001_011",
                       m22500_24999 = "B20001_012",
                      m25000_29999 = "B20001_013",
                      m30000_34999 = "B20001_014",
                       m35000_39999 = "B20001_015",
                       m40000_44999 = "B20001_016",
                       m45000_49999 = "B20001_017",
                       m50000_54999 = "B20001_018",
                       m55000_64999 = "B20001_019",
                       m65000_74999 = "B20001_020",
                       m75000_99999 = "B20001_021",
                       m100000_ ="B20001_022",
                       f1_2499 =  "B20001_024",
                       f2500_4999 = "B20001_025",
                       f5000_7499 = "B20001_026",
                       f7500_9999 = "B20001_027",
                       f10000_12499 = "B20001_028",
                       f12500_14999 = "B20001_029",
                       f15000_17499 = "B20001_030",
                       f17500_19999 = "B20001_031",
                       f20000_22499 = "B20001_032",
                       f22500_24999 = "B20001_033",
                       f25000_29999 =  "B20001_034",
                       f30000_34999 = "B20001_035",
                       f35000_39999 = "B20001_036",
                       f40000_44999 = "B20001_037",
                       f45000_49999 = "B20001_038",
                       f50000_54999 = "B20001_039",
                       f55000_64999 = "B20001_040",
                       f65000_74999 = "B20001_041",
                      f75000_99999 = "B20001_042",
                      f100000_ = "B20001_043"), state=state, survey="acs5"))
}

dc_pay <- load_data_pay("DC")
md_pay <- load_data_pay("MD")
va_pay <- load_data_pay("VA")

dmv_pay <- rbind(dc_pay, md_pay, va_pay)
dmv_pay <- dmv_pay[,-5]
dmv_pay2 <- dmv_pay %>% spread(variable, estimate) #transpose the male and female variables

print(head(dmv_pay2))


```

### Clean data now

We want to clean the above data and calculate the total sum of all female income, total sum of all male income, and then the difference between averages. 

```{r cleaning, echo=TRUE}

names <- names(dmv_pay2)[3:length(names(dmv_pay2))]
maths <- ifelse(names %in% c("f100000_","m100000_"), 200000 , gsub("f","", gsub("m", "", gsub("_", "+", names))))
maths2 <- sapply(maths, FUN=function(x) (eval(parse(text=x))+1)/2) # we will apply the average of each group to each column

malesum = rep(0, nrow(dmv_pay2))
femalesum = rep(0, nrow(dmv_pay2))
for(i in 1:length(names)){
  if(substr(names[i],1,1)=="f"){
    femalesum <- femalesum + dmv_pay2[[names[i]]]*as.numeric(maths2[i])
  }
  else {
    malesum <- malesum + dmv_pay2[[names[i]]]*as.numeric(maths2[i])
  }
}

dmv_pay2$malesum <- malesum
dmv_pay2$femalesum <- femalesum

## merge dmv_fm2 to get female and male totals
dmv_pay3 <- inner_join(dmv_pay2, dmv_fm2) #should be the same number of rows
if(nrow(dmv_pay3)!= nrow(dmv_pay2)) print("ERROR CHECK OUT")

dmv_pay3$femaleavg <- round(dmv_pay3$femalesum/dmv_pay3$female,2)
dmv_pay3$maleavg <- round(dmv_pay3$malesum/dmv_pay3$male,2)

dmv_pay3$dif <- dmv_pay3$femaleavg - dmv_pay3$maleavg #difference between female and male
#positive is better

print(dmv_pay3[, c(1:2, 44:length(names(dmv_pay3)))])
```

### Comparing the Pay Gap by Count

Now we will map our country differences. 

```{r mapavg, echo=TRUE}
dmv_pay4 <- geo_join(dmv, dmv_pay3, "GEOID", "GEOID", how="inner") #158 rows
if(nrow(dmv_pay4@data) != nrow(dmv_pay3)) print("ERROR CHECK OUT")

pal2 <- colorBin("Blues", domain=c(min(dmv_pay3$dif), max(dmv_pay3$dif)), bins = 8)

map2 <- leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = dmv_pay4, 
              fillColor = ~pal2(dmv_pay4$dif), 
              fillOpacity = 0.7, 
              weight = 0.2, 
              smoothFactor = 0.2, 
              popup = paste("Difference:", format(dmv_pay4$dif, big.mark=","),sep=" ")) %>%
  addLegend(pal = pal2, 
            values = dmv_pay4$dif, 
            position = "bottomright", 
            title = "Income Difference Men - Women",
            labFormat = labelFormat(prefix="$")) #%>% labFormat("bin")

map2
```

### Examine the Average Amount of Differences

```{r avgs, echo=TRUE}
library(scales)
differences <- dmv_pay3[,c(1:2, 48:50)]
differences2 <- differences[,-5] %>% gather(gender, amount, femaleavg:maleavg, factor_key=TRUE)

diffavg <- differences2 %>% group_by(gender) %>% summarize(mean.inc = mean(amount))

ggplot(data=differences2, aes(x=amount, fill=gender,color=gender)) + geom_histogram(aes(y=..density..),alpha=.5) +
  geom_density(alpha=.2) + geom_vline(data=diffavg, aes(xintercept=mean.inc,  colour=gender),
               linetype="dashed", size=1) + ggtitle("Female and Male Average Comparisons") +
  ylab("Density") + xlab("Average Income by County") + theme_bw() + 
  scale_y_continuous(labels = scales::percent) + scale_x_continuous(labels = scales::comma)
```

```{r examine, echo=TRUE}

fem.higher = length(which(dmv_pay3$dif>0))
total = length(dmv_pay3$dif)

print(paste("There are only", fem.higher, "counties, out of", total, "where females make more than men"))

print(dmv_pay3[dmv_pay3$dif>0,c(2, 43:50)])

```


## Citations
D. Kahle and H. Wickham. ggmap: Spatial Visualization with ggplot2. The R Journal, 5(1),
  144-161. URL http://journal.r-project.org/archive/2013-1/kahle-wickham.pdf